---
title: "Heimadæmi 4"
author: "Signý Kristín Sigurjónsdóttir"
date: "`r Sys.Date()`"
output: 
    html_document:
        code_folding: hide
        theme: flatly
        toc: true
        toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.asp = 0.621, fig.width = 8, out.width = "100%")
```

```{r warning=F, message=F, echo=F}
library(tidyverse)
library(rmdformats)
library(grid)
library(gridExtra)
library(rethinking)
library(dagitty)
library(kableExtra)
```

# 13. kafli

## 13E1
Which of the following priors will produce more shrinkage in the estimates?

(a) $\alpha_{tank} \sim \mathrm{Normal}(0,1)$;

(b) $\alpha_{tank} \sim \mathrm{Normal}(0,2)$.

***

**Lausn**: Prófum að teikna þessa prior-a upp:

```{r}
ggplot(data=data.frame(x=-4:4), aes(x)) + stat_function(fun=dnorm, n=1000, args = list(mean = 0, sd = 1), size=1.2, aes(color="a")) + stat_function(fun=dnorm, n=1000, args = list(mean = 0, sd = 2), size=1.2, aes(color="b")) + scale_colour_manual("Legend", values = c("dodgerblue", "indianred3"))
```
Sjáum að *a* gefur meiri rýrnun (e. shrinkage), en það er vegna þess að *a* er með minna staðalfrávik.

## 13E2
Rewrite the following model as a multilevel model.

$$
y_i \sim \mathrm{Binomial}(1,p_i) \\
\mathrm{logit}(p_i) = \alpha_{group[i]} + \beta x_i \\
\alpha_{group} \sim \mathrm{Normal}(0,1.5) \\
\beta \sim \mathrm{Normal}(0,0.5)
$$

***

**Lausn**: Til þess að framkvæma umbeðnar breytingar þarf að breyta hvernig $\alpha_{group}$ er skilgreint og bæta við tveimur priors. Sjáum nýja módelið (multilevel) hér að neðan.

$$
y_i \sim \mathrm{Binomial}(1,p_i) \\
\mathrm{logit}(p_i) = \alpha_{group[i]} + \beta x_i \\
\alpha_{group} \sim \mathrm{Normal}(\bar{\alpha}, \sigma_{\alpha}) \\
\beta \sim \mathrm{Normal}(0,0.5) \\
\bar{\alpha} ∼ \mathrm{Normal}(0, 1.5) \\
\sigma_{\alpha} ∼ \mathrm{Exponential}(1)
$$

## 13E3
Rewrite the following model as a multilevel model.

$$
y_i \sim \mathrm{Normal}(\mu_i,\sigma) \\
\mu_i = \alpha_{group[i]} + \beta x_i \\
\alpha_{group} \sim \mathrm{Normal}(0,5) \\
\beta \sim \mathrm{Normal}(0,1) \\
\sigma \sim \mathrm{Exponential}(1)
$$

***

**Lausn**: Líkt og í dæminu á undan fæst:



$$
y_i \sim \mathrm{Normal}(\mu_i,\sigma) \\
\mu_i = \alpha_{group[i]} + \beta x_i \\
\alpha_{group} \sim \mathrm{Normal}(0,5) \\
\beta \sim \mathrm{Normal}(0,1) \\
\sigma \sim \mathrm{Exponential}(1) \\
\bar{\alpha} ∼ \mathrm{Normal}(0, 1.5) \\
\sigma_{\alpha} ∼ \mathrm{Exponential}(1)
$$

## 13E4
Write a mathematical model formula for a Poisson regression with varying intercepts.

***

**Lausn**: Styðjumst við módelin í **13E2** og **13E3.**

$$
y_i \sim \mathrm{Poisson}(\lambda_i) \\
\mathrm{log}(\lambda_i) = \alpha_{group[i]} + \beta x_i \\
\alpha_{group} \sim \mathrm{Normal}(\bar{\alpha}, \sigma_{\alpha}) \\
\beta \sim \mathrm{Normal}(0,0.5) \\
\bar{\alpha} ∼ \mathrm{Normal}(0, 1.5) \\
\sigma_{\alpha} ∼ \mathrm{Exponential}(1)
$$






## 13E5
Write a mathematical model formula for a Poisson regression with two different kinds of varying
intercepts, a cross-classified model.

**Lausn:**

***


# 14. kafli



## 14E1
Add to the following model varying slopes on the predictor x.
$$
y_i \sim \mathrm{Normal}(\mu_i,\sigma) \\
\mu_i = \alpha_{group[i]} + \beta x_i \\
\alpha_{group} \sim \mathrm{Normal}(\alpha,\sigma_{\alpha}) \\
\alpha \sim \mathrm{Normal}(0,10) \\
\beta \sim \mathrm{Normal}(0,1) \\
\sigma \sim \mathrm{Exponential}(1) \\
\sigma_{\alpha} \sim \mathrm{Exponential}(1) 
$$

**Lausn:**

***


## 14E2
Think up a context in which varying intercepts will be positively correlated with varying
slopes. Provide a mechanistic explanation for the correlation.

**Lausn:**

***


## 14E3
When is it possible for a varying slopes model to have fewer effective parameters (as estimated
by WAIC or PSIS) than the corresponding model with fixed (unpooled) slopes? Explain.

**Lausn:**

***


